---
title: "Geography PhDs"
author: "Laura Sheppard"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Geography PhDs in the UK

This markdown file and R code accompanies the article `Finding the female geographers: The gendered dynamics of UK geography PhD study` (Sheppard and Reades, 2024) in `The Professional Geographer`.

The source data used to create this database of Geography PhDs is the April 2023 version of the British Library's E-Thesis Online Service (EThOS) metadata, where there are 627,042 doctoral records and 15 variables. (https://bl.iro.bl.uk/concern/datasets/10cc13f9-797d-41f2-a7e2-d29f4306133e)

The variables in EThOS include:
  + Thesis title
  + DOI
  + Author name
  + Author ISNI
  + ORCID
  + Institution
  + Institutional ISNI
  + Date of award
  + Qualification
  + Abstract
  + Subject Discipline
  + Supervisor
  + Funder(s)
  + EThOS URL
    
Outline of the markdown file:
  + Cleaning the data
  + Identifying geography PhDs
  + Combining geography PhDs
  + Removing duplicate records
  + Finalising PhD database

### Scope and clean the EThOS data

```{r packages, echo = FALSE, include = FALSE}
#load relevant packages
library(dplyr)
library(tidyr)
library(tidyverse)
library(janitor)
library(stringi)
library(stringr)
library(ggplot2)
library(readr)
library(tm)
library(mosaic)
```

```{r read in the EThOS data, echo = FALSE, include = FALSE}
EThOS <- read_csv("EThOS_CSV_202304.csv") %>%
  clean_names() #clean column names using janitor

#https://bl.iro.bl.uk/concern/datasets/37e3752f-5360-41d4-9d8c-b54dcabb9d1c
#April 2023 - 627,042 records, 15 variables
```

```{r extract EThOS IDs, echo = FALSE, include = FALSE}
#Extract EThOS IDs from EThOS URLs
EThOS <- EThOS %>%
  separate(e_th_os_url, c('EThOS_URL', 'EThOS_ID'), sep = 'bl.ethos') 

#Remove remaining full stop
EThOS$EThOS_ID <- gsub('\\.', '', EThOS$EThOS_ID)
```

```{r extract ISNI IDs, echo = FALSE, include = FALSE}
#Extract HEIs ISNI from institution isni URLs
EThOS <- EThOS %>%
  separate(institution_isni, c('institution_URL', 'institution_isni'), sep = '/isni/') 

#Remove remaining full stop
EThOS$institution_isni <- gsub('\\.', '', EThOS$institution_isni)
```

```{r select columns needed, echo = FALSE, include = FALSE}
#select only the necessary columns
EThOS <- EThOS %>%
  dplyr::select(title, author, author_isni, institution, institution_isni, date, qualification, subject_discipline, supervisor_s, abstract, EThOS_ID, ir_url)
```

For the PhD research, the EThOS team at the BL provided additional data that is not publicly available. 

```{r add in the BL's additional data, echo = FALSE, include = FALSE}
#additional EThOS data provided by the BL
EThOS_AdditionalData <- read_csv("EThOS_CSV_202304_AdditionalData.csv") %>%
  clean_names()

#change the column names
EThOS_AdditionalData <- rename(EThOS_AdditionalData, c(EThOS_ID = "x1")) 
EThOS_AdditionalData <- rename(EThOS_AdditionalData, c(ddc_number = "x2"))
EThOS_AdditionalData <- rename(EThOS_AdditionalData, c(keywords = "x4"))
EThOS_AdditionalData <- rename(EThOS_AdditionalData, c(department = "x5"))

#as numeric
EThOS_AdditionalData$ddc_number <- as.numeric(EThOS_AdditionalData$ddc_number)
```

```{r join additional EThOS data to the main EThOS data, echo = FALSE, include = FALSE}
#add additional data to main EThOS data
#as numeric
EThOS$EThOS_ID <- as.numeric(EThOS$EThOS_ID)

EThOS <- EThOS %>%
  left_join(.,
            EThOS_AdditionalData, 
            by = c("EThOS_ID" = "EThOS_ID"))

EThOS <- EThOS %>%
  distinct(title, institution, EThOS_ID, .keep_all = TRUE)
```

```{r add in DDC data, echo = FALSE, include = FALSE}
#Dewey Decimal Classifications
#extract numbers and labels to merge

EThOS_DDC <- read_csv("DDC.csv") %>%
  clean_names() %>%
  group_by(ddc, ddc2) %>%
  summarise(ddc_total = n())

#truncate ddcs
EThOS$ddc_number <- trunc(EThOS$ddc_number*10^2)/10^2 #to 2 decimal places - easier for data matching
EThOS_DDC$ddc <- trunc(EThOS_DDC$ddc*10^2)/10^2 #to 2 decimal places - easier for data matching
```

```{r join to DDC data to main EThOS data}
#as numeric
EThOS_DDC$ddc <- as.numeric(EThOS_DDC$ddc)

EThOS <- EThOS %>%
  left_join(.,
            EThOS_DDC, 
            by = c("ddc_number" = "ddc"))

EThOS <- EThOS %>%
  distinct(title, institution, EThOS_ID, .keep_all = TRUE)
```

```{r filter by qualification type}

#filter for PhDs rather than other qualifications 
EThOS <- EThOS %>%
  filter(qualification == "Thesis (Ph.D.)")

#removes around 37,000 records, leaves 590,900 records
```

```{r remove duplicate column and rename}
EThOS$subject_discipline.y <- NULL

EThOS <- EThOS %>%
  rename(., subject_discipline = "subject_discipline.x")
```

### Filter / identify geography PhD theses

As the EThOS data has been cleaned and additional data has been added in, we can now work on identifying geography PhDs.

Four ways:
+ Geography and travel sub-discipline 
+ Geography as a key word
+ Geography in the title of the thesis
+ Geography departments

```{r identify geography PhDs - DDC}
#how to identify geography PhD thesis
#change to lower case
EThOS$title <- tolower(EThOS$title)
EThOS$keywords <- tolower(EThOS$keywords)
EThOS$department <- tolower(EThOS$department)

#sub discipline of geography and travel
geography_sub_discipline <- EThOS %>%
  filter(ddc2 == "Geography & travel")
```

```{r other ways to identify geography PhDs}
geography_dep <- EThOS %>%
  filter(grepl("geography|geograph|geospatial|geographical|geographic|cartographic", department))

geography_titles <- EThOS %>%
  filter(grepl("geography|geograph|geospatial|geographical|geographic|cartographic", title))

geography_keywords <- EThOS %>%
  filter(grepl("geography|geograph|geospatial|geographical|geographic|cartographic", keywords))
```

Filtering for the word geography in the title, key word, and department columns finds more "geography" theses than simply selecting on the sub-discipline of geography and travel. Less theses contain geography as a key word than the other two categories. 

+ Key words - 2392 PhDs
+ Sub discipline - 1599 PhDs
+ Titles - 1857 PhDs
+ Departments - 3263 PhDs

```{r Geography PhDs over time 1}
#Using KEY WORDS

Geog_years1 <- geography_keywords %>%
  group_by(date) %>%
  summarise(count_years = n()) %>%
  filter(date >= 1917) %>%
  filter(date <= 2022)
  
ggplot(Geog_years1, 
       aes(x=date)) + 
  geom_line(aes(y = count_years, colour = "PhD students")) + 
  scale_colour_manual(values=c(`PhD students` ="Dark Blue")) +
  theme(legend.position="top") +
  labs(title = paste0("Number of PhD students in Geography (using key words)"),
       subtitle = paste0("", length(unique(geography_keywords$institution)), " Higher Education Institutions"),
       color = "Number of Geography PhD students",
       y = "Number of PhDs per year",
       x = "Year the thesis was awarded")
```

```{r Geography PhDs over time 2}
#Using TITLES

Geog_years2 <- geography_titles %>%
  group_by(date) %>%
  summarise(count_years = n()) %>%
  filter(date >= 1917) %>%
  filter(date <= 2022)
  
ggplot(Geog_years2, 
       aes(x=date)) + 
  geom_line(aes(y = count_years, colour = "PhD students")) + 
  scale_colour_manual(values=c(`PhD students` ="Dark Blue")) +
  theme(legend.position="top") +
  labs(title = paste0("Number of PhD students in Geography (using titles)"),
       subtitle = paste0("", length(unique(geography_titles$institution)), " Higher Education Institutions"),
       color = "Number of Geography PhD students",
       y = "Number of PhDs per year",
       x = "Year the thesis was awarded")
```

```{r Geography PhDs over time 3}
#Using SUB DISCIPLINES

Geog_years3 <- geography_sub_discipline %>%
  group_by(date) %>%
  summarise(count_years = n()) %>%
  filter(date >= 1917) %>%
  filter(date <= 2022)
  
ggplot(Geog_years3, 
       aes(x=date)) + 
  geom_line(aes(y = count_years, colour = "PhD students")) + 
  scale_colour_manual(values=c(`PhD students` ="Dark Blue")) +
  theme(legend.position="top") +
  labs(title = paste0("Number of PhD students in Geography (using ddc numbers)"),
       subtitle = paste0("", length(unique(geography_sub_discipline$institution)), " Higher Education Institutions"),
       color = "Number of Geography PhD students",
       y = "Number of PhDs per year",
       x = "Year the thesis was awarded")
```

```{r Geography PhDs over time 4}
#Using DEPARTMENTS

Geog_years4 <- geography_dep %>%
  group_by(date) %>%
  summarise(count_years = n()) %>%
  filter(date >= 1917) %>%
  filter(date <= 2022)
  
ggplot(Geog_years4, 
       aes(x=date)) + 
  geom_line(aes(y = count_years, colour = "PhD students")) + 
  scale_colour_manual(values=c(`PhD students` ="Dark Blue")) +
  theme(legend.position="top") +
  labs(title = paste0("Number of PhD students in Geography (using departments)"),
       subtitle = paste0("", length(unique(geography_keywords$institution)), " Higher Education Institutions"),
       color = "Number of Geography PhD students",
       y = "Number of PhDs per year",
       x = "Year the thesis was awarded")
```

I decided to combine all five ways of identifying which PhDs are geography ones in order to "find" all possible geography PhD theses. I then used the distinct function from dplyr so that there were not duplicates included in the final geography PhD data set.

```{r combining geography PhDs}
#combine three together and use distinct
geography_PhDs <- rbind(geography_keywords, geography_sub_discipline, geography_titles, geography_dep) #, geography_abstract)

geography_PhDs <- geography_PhDs %>%
  distinct(title, institution, EThOS_ID, .keep_all = TRUE)
```

+ 7,425 geography PhDs (after distinct)

```{r Geography PhDs over time 5}
#Using ALL FOUR METHODS

Geog_years5 <- geography_PhDs %>%
  group_by(date) %>%
  summarise(count_years = n()) %>%
  filter(date >= 1917) %>%
  filter(date <= 2022)
  
ggplot(Geog_years5, 
       aes(x=date)) + 
  geom_line(aes(y = count_years), linetype = "solid") + 
  labs(title = paste0("Geography PhD students over time"),
       subtitle = paste0("", length(unique(geography_PhDs$institution)), " Higher Education Institutions"),
       color = "Number of Geography PhD students",
       y = "Number of Geography PhDs each year",
       x = "Awarding year") +
  theme_minimal() +
  theme(legend.position = "none")

ggsave("Geography_PhDs_1917_to_2023_line.png")
```

```{r}
write_csv(geography_PhDs, "Geography_PhDs.csv")
```

